Deciding Bitvector Termination with Program Synthesis

### 2014-07-06 papers/termination/paper.tex
various edits

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
\emph{ranking function} for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of ranking functions is
insufficient for proving termination of many terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.

@@

 For this purpose, we phrase the termination and non-termination problems as
second-order satisfaction problems.  We identify a fragment of second-order
logic that is expressive enough to capture our formulation, which we call
the \emph{synthesis fragment}.  Subsequently, we propose a method for
checking the satisfiability of a formula in the synthesis fragment by
solving an isomorphic program synthesis problem
(Section~\ref{sec:synthesis}).  Our method is sound and semi-complete, i.e. 
if a pair of a ranking function and a supporting invariant exist it will be
found, but if the program does not terminate we may not find a recurrence
set.  However if the program's transition relation is deterministic, our
method is complete.

With respect to the performance of our technique, we show that its runtime is dominated by the size of 
the shortest termination proof.
%While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors 
%influencing our method's performance and
%find that our technique  
%
This means that our procedure is not directly dependent on the structure of
the analysed loop, e.g.  number of variable, number of lassos, but on the
Kolmogorov complexity of its minimal termination argument.  We argue that
the length of a program's termination proof is indicative of how easy to
understand the program is, and thus programmers tend to write programs that
have relatively short termination proofs.  We empirically show that this
claim holds in practice, causing our technique to perform well.

@@

\section{Motivational Examples} \label{sec:motivation}
%- see phase change programs.

Figure~\ref{fig:handletable} illustrates the most common simplifying
assumptions made by existing termination analyses:
%
\begin{itemize}
\item[(i)] programs use only linear arithmetic.
\item[(ii)] terminating programs have termination arguments in linear arithmetic.
\item[(iii)] the semantics of bit-vectors and mathematical integers are equivalent.
\item[(iv)] the semantics of floats and mathematical reals are equivalent.
\end{itemize}

%In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}).
To show how these assumptions are broken by even simple programs, we draw the reader's attention to the programs in
Figure~\ref{fig:motivation} and their curious properties:

%We next discus some intricate examples that prove challenging for existent termination analysis.

%\subsection{Program linearity vs. the linearity of its ranking function.}

%Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.

\begin{itemize}

\item Program (a) breaks assumption (i) as it makes use of the bit-wise $\&$ operator.
%As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
%although a linear ranking function does exist (see Figure~\ref{fig:handletable}). 
%In order to find a ranking function for this example, it is necessary to take into account
%the semantics of the bit-wise AND operator.
Our technique finds that an admissible ranking function is the linear function
$R(x) = x$, whose value decreases with
every iteration, but cannot decrease indefinitely as it is bounded from below.
This example also illustrates the lack of a direct correlation between the linearity of a program and that of its termination arguments.

\item Program (b) break assumption (ii), in that it has no linear ranking function.
We prove that this loop terminates by finding the non-linear ranking function $R(x) = |x|$.

@@

\item Program (d) also breaks assumption(iii), but ``the other way'' -- it terminates for integers but not for bit-vectors.
If each of the variables is stored in an unsigned $k$-bit word, the following configuration will
lead to an infinite loop:
$$M = 2^k - 1, N = 2^k - 1, i = M, j = N-1$$

\item Program (e) breaks assumption (iv) -- it terminates for reals but not for floats.
If $x$ is sufficiently large, rounding error will cause the subtraction to have no effect.

\item Program (f) breaks assumption (iv) ``the other way'' -- it terminates for floats but not for reals.
Eventually $x$ will become sufficiently small that the nearest representable number is 0.0, at which point
it will be rounded to 0.0 and the loop will terminate.
\end{itemize}

Up until this point, we looked at examples that are not soundly treated by existing techniques as they 
don't fit in the range of programs covered by these techniques.
Next, we look at some programs that are handled by existing termination tools via dedicated analyses. We show that our method 
uniformly handles them, without the need for any special treatment.
% we are able to compute termination arguments in several interesting case

% Our technique gains from generality. A possible explanation for this (perhaps initially sur-
%prising) fact is that, for synthesis, we are interested in the mere existence of a
%% solution, and the loss of
%% many solutions does not necessarily mean the loss of
%% all solutions of the constraint
% bit-level analysis
%These non-terminating programs represent an important class of
%bugs, especially for denial-of-service related vulnerabilities, where
%an attacker can exploit the overflow. Equally important,
% equally important

\begin{itemize}
\item Program (g) is a linear program
which is shown in~\cite{DBLP:conf/tacas/CookSZ13} not to admit (without prior manipulation) a lexicographic linear function.
With our technique we can find the non-linear ranking function $R(x) = |x|$.
%As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
%loop transition leads to a decrease with respect to the lexicographic
%ordering for this tuple. Therefore no function may increase unless a function of
%a lower index decreases. Additionally, at every step, there must be at least one
%function that decreases.

@@

\item In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i)
admits a multiphase ranking function computed from a multiphase ranking
template.  A multiphase ranking template is targeted at programs that go
through a finite number of phases in their execution.  Each phase is ranked
with an affine-linear function and the phase is considered to be completed
once this function becomes non-positive.

In our setting this type of programs does not need special treatment, as we
can find a non-linear lexicographic ranking function $R(x, y, z) = (x < y,
z)$.  This termination argument is somewhat subtle, but can be understood as
follows: the boolean quantity $x < y$ eventually decreases, that is to say
once a state with $x \geq y$ is reached, $x$ never again becomes greater
than $y$.  This means that as soon as the ``else'' branch of the if
statement is taken, it will continue to be taken in each subsequent
iteration of the loop.  Meanwhile, if $x < y$ has not decreased (i.e.  we
have stayed in the same branch of the ``if'') then $z$ does decrease.  Since
a boolean only has two possible values, it cannot decrease indefinitely. 
Since $z > 0$ is a conjunct of the loop guard, $z$ cannot decrease
indefinitely, and so $R$ proves that the loop is well founded.

As with all of the termination proofs presented in this paper, the ranking
function above was found completely automatically.

\iffalse
\item Program (j) illustrates a nested loop.  This construction can cause difficulties for provers based on
enumerating lassos
\fi
\end{itemize}

@@

\section{Preliminaries}
Given a program, %with state space $X$ and transition relation $T \subseteq X \times X$, 
we first formalise its termination argument as a ranking function
(Section~\ref{sec:ranking.functions}).  Subsequently, we discus bit-vector
semantics and illustrate differences between machine arithmetic and integer
arithmetic that make the abstraction of bit-vectors to mathematical integers
unsound (Section~\ref{sec:machine.arith}).

\subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
A programs $P$ is represented as a transition systems with state space $X$, transition relation $T \subseteq X \times X$, and set of variables $x$. 
A state of the program is a valuation of the variables from $x$. For a state $s \in X$ with $T(s,s')$ we say $s'$
is the successor of $s$ under $T$.

\begin{definition}[Unconditional termination]
A transition system is said to be \emph{unconditionally terminating} if there is no infinite sequence of states
$x_1, x_2, \ldots \in X$ with $\forall i . T(x_i, x_{i+1})$.  
\end{definition}

We can prove that the transition system is unconditionally terminating by finding a ranking function for its transition relation.

\begin{definition}[Ranking function]
An injective function ${R:X\to Y}$ is a \emph{ranking function} for the transition relation $T$ if $Y$ is a well-founded set and 
$R$ is monotonically decreasing with respect to $T$.  That is
to say:
$$\forall x, x' \in X. T(x, x') \Rightarrow R(x) < R(x')$$
\end{definition}

@@

In the case that $\dim(Y) = 1$, this reduces to an inner product:
$$f(\vec{x}) = \vec{\lambda} \cdotp \vec{x} + c$$

@@

\subsection{Machine Arithmetic and Bitvectors} \label{sec:machine.arith}

Physical computers have bounded storage, which means they are unable to
perform calculations on mathematical integers.  For example, if $A$ is the
Ackermann function and $G$ is Graham's number, a physical computer capable
of computing $A(G, G)$ would contain (much!) more matter than is believed to
exist in the universe.  Fortunately, it is rare for a programmer to need
such a large number and so modern computers do their arithmetic over
fixed-width binary words, otherwise known as bit-vectors.  For the remainder
of this section, we will say that the bit-vectors we are working with are
$k$-bits wide, which means that each word can hold one of $2^k$ bit
patterns.  Typical values for $k$ are 32 and 64.

Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
of the differences between machine arithmetic and Peano arithmetic.  To give
an example: $(2^k - 1) + 1 \equiv 0 \pmod {2^k}$.  This provides a
counterexample to the statement $\forall x .  x + 1 > x$, which is a theorem
of Peano arithmetic but not of modular arithmetic.  When an arithmetic
operation has a result greater than $2^k$, it is said to ``overflow''.  If
an operation does not overflow, its machine-arithmetic result is the same as
the result of the same operation performed on integers.

Machine words can be interpreted as ``signed'' or ``unsigned'' values. 
Signed values can be negative, while unsigned values cannot.  The encoding
for signed values is two's complement, where the most significant bit
$b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather
than $2^k - 1$.  Two's complement representation has the property that
$\forall x .  -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
is bitwise-negation.  Two's complement also has the property that addition,
multiplication and subtraction are defined identically for unsigned and
signed numbers.  Therefore signed and unsigned arithmetic differ only on
comparison operators\footnote{There is also the difference that signed
overflow in a C program results in undefined behaviour, but in practice the
undefined behaviour is implemented just as if the arithmetic had been
unsigned}.  For example: $$\forall x .  -x \leq_s x \wedge -x \geq_u x$$
where $\leq_s$ and $\geq_u$ represent signed and unsigned comparisons
respectively.

The final source of difference between integer arithmetic and bitvector
arithmetic stems from width conversions.  Typical computers allow variables
to have different types, which can be represented using words of different
widths.  In C a \texttt{short} might occupy 16 bits, while an \texttt{int}
might occupy 32 bits.  If a $k$-bit variable is assigned to a $l$-bit
variable with $l < k$, the result is truncated $\mod 2^l$.  For example, if
$x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will hold the
value $0$ after the following code is executed:

\begin{lstlisting}
x = 65536;
y = x;
\end{lstlisting}

@@

As well as machine arithmetic differing from Peano arithmetic on the
operators they have in common, computers have several ``bitwise'' operations
that are not taken as primitive in the theory of integers.  These operations
include the standard boolean operations \texttt{and, or, not, xor} applied
to each element of the bit-vector.  Computer programs often make use of
these operators, which are non-linear when interpreted in the standard model
of Peano arithmetic\footnote{ Although some of these operators can be seen
as linear in a different algebraic structure, e.g.  \texttt{xor} corresponds
to addition in the Galois field $\mathrm{GF}(2^k)$}.

@@

\begin{definition}[Synthesis Fragment]
 A formula is in the synthesis fragment iff it is of the form
 \[
  \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} . \sigma(\vec{P}, \vec{x}, \vec{y})
 \]

where $\vec{P}$ ranges over sets, while $\vec{x}$ and $\vec{y}$ range over ground terms.
Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to \mathbb{B}$ can be viewed as a specification function
that  returns true iff the functions $\vec{P}$ compute appropriate outputs
when fed the inputs $\vec{x}$ and $\vec{y}$.  
\end{definition}
%
Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
%Solving the synthesis problem 
amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
for all $\vec{y}$. 
If a pair $(\vec{P_0}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P_0}, \vec{x}) \models \sigma$.
For the remainder of the presentation, we drop the $\vec{x}$ and instead just write $x$, with the understanding
that all variables range over vectors.

@@

We will begin our discussion by showing how to encode the (non-)termination
of program consisting of a single loop in the synthesis fragment.  For the
time being, a loop is a pair ${L= \langle G,T \rangle}$, where the states
$x$ satisfying the loop's guard are given by the predicate $G(x)$.  The body
of the loop is encoded as the transition relation $T(x, x')$, meaning that
state $x'$ is reachable from state $x$ via a single iteration of the loop
body.  For example, the loop in Figure~\ref{fig:motivation.a} is encoded as:
%
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

\begin{definition}[Non-Termination Formula - Open Recurrence Set  {\bf [ONT]}]
\label{def:ont}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T_u(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula - Closed Recurrence Set {\bf [CNT]}]
\label{def:cnt}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T_u(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

As the existence of a ranking function is equivalent to the satisfiability
of the formula {\bf [UT]}, a satisfiality witness 
is also a ranking function and thus a proof of $L$'s unconditional termination.

Continuing the example of the program in Figure~\ref{fig:motivation.a}, we can see that
the corresponding synthesis formula {\bf [UT]} is satisfiable, as witnessed by the function $R(x) = x$.
So $R(x) = x$ constitutes a proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
terminating.

Note that different formulations for unconditional termination are possible.
We are aware of a proof rule based on transition invariants, i.e.  supersets
of the transition relation's transitive closure
\cite{DBLP:conf/pldi/GrebenshchikovLPR12}.  This formulation assumes that
the second-order logic has a primitive predicate for disjunctive
well-foundedness.  Conversely, our formulation in Definition~\ref{def:UT}
does not require a transitive closure operator, nor a disjunctive
well-foundedness predicate.  \\

@@

Gupta et al. \cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination of a transition relation $T$ by the existence of an \emph{(open) recurrence set},
i.e. a nonempty set of states $N$ such that for each $s \in N$ there
exists a transition to some $s'\in N$.
For our case, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.
Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition relation, $T_u$. 
%$L = \langle G,T \rangle$
Again, this distinction is mostly important when dealing with nested loops, as illustrated in Section~\ref{sec:env}.

@@

If this formula is satisfiable, $N$ is an open recurrence set for $L$ which proves
$L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to the synthesis fragment
(it is an $\exists \forall \exists$ formula). %This means that we will not be able to use the solver for the synthesis fragment to solve it.

In order to formulate the non-termination problem inside the synthesis fragment, 
we make use of the notion of \emph{closed recurrence set} introduced by Chen et al. in \cite{DBLP:conf/tacas/ChenCFNO14}. 
As opposed to an open recurrence set, this is a $\exists \forall$ property: for each state in the recurrence set $N$, all of its successors 
must be in $N$. We note that if $T$ is deterministic, every open recurrence set is also a closed recurrence set (since each
state has at most one successor).  Thus, the non-termination problem for deterministic transition systems is equivalent to the
satisfiability of formula {\bf [CNT]}. %Definition~\ref{def:deterministic-nonterm-formula}.
%As our formulation is with respect to a given loop $L=\langle I,G,T \rangle$, we require the  
%recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$. In other words, the loop's guard needs to be satisfied in order for the 
%infinite execution to happen.

@@

Note that by defining $T_u$'s determinism as the existence of exactly one successor for any state satisfying $L$'s guard, 
we eliminate the possibility of $T_u$ being $false$, which would otherwise constitute a counter-example for Theorem~\ref{thm:cnt}.

Given that our program's state space is finite, a transition relation induces a infinite execution if and only if some state is 
visited infinitely often, or equivalently the state is visited twice in the course of the execution.
This gives us an alternative way of expression a non-termination proof: we can can try to find a witness $x$ to the satisfiability
of the formula
$$ \exists x . T^+(x, x)$$
However, deciding satisfiability of this formula directly requires a logic that includes a transitive closure operator, which
ours does not have.\\

\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based on checking the satisfiability of formula {\bf [CNT]}
may fail to find a closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is non-deterministic.  
To understand the issue, imagine the following loop with an non-deterministic assignment:

>>>>

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
\emph{ranking function} for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of ranking functions is
insufficient for prove termination for many terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.

@@

 For this purpose, we phrase the termination and non-termination problems as
second-order satisfaction problems.  We identify a fragment of second-order
logic that is expressive enough to capture our formulation, which we call
the \emph{synthesis fragment}.  Subsequently, we propose a method for
checking the satisfiability of a formula in the synthesis fragment by
solving an isomorphic program synthesis problem
(Section~\ref{sec:synthesis}).  Our method is sound and semi-complete, i.e. 
if a pair of a ranking function and a supporting invariant exist it will be
found, but if the program does not terminate we may not find a recurrence
set.  However, if the program's transition relation is deterministic, our
method is complete.

With respect to the performance of our technique, we show that its runtime is dominated by the size of 
the shortest termination proof.
%While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors 
%influencing our method's performance and
%find that our technique  
%
This means that our procedure is not directly dependent on the structure of
the analysed loop (e.g.~the number of variables or the number of lassos),
but on the \emph{Kolmogorov complexity of its minimal termination argument}. 
We argue that the length of a program's termination proof is indicative of
how easy to understand the program is, and thus programmers tend to write
programs that have relatively short termination proofs.  We empirically show
that this claim holds in practice, causing our technique to perform well.

@@

\section{Motivating Examples} \label{sec:motivation}
%- see phase change programs.

Figure~\ref{fig:handletable} illustrates the most common simplifying
assumptions made by existing termination analyses:
%
\begin{itemize}
\item[(i)] programs use only linear arithmetic.
\item[(ii)] terminating programs have termination arguments expressible in linear arithmetic.
\item[(iii)] the semantics of bit vectors and mathematical integers are equivalent.
\item[(iv)] the semantics of IEEE floating-point numbers and mathematical reals are equivalent.
\end{itemize}

%In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}).
%
To show how these assumptions are violated by even simple programs, we draw
the reader's attention to the programs in Figure~\ref{fig:motivation} and
their curious properties:
%
%We next discus some intricate examples that prove challenging for existent termination analysis.
%
%\subsection{Program linearity vs. the linearity of its ranking function.}
%
%Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.
%
\begin{itemize}

\item Program (a) breaks assumption (i) as it makes use of the bit-wise $\&$ operator.
%
%As such, Figure~\ref{} denotes a program with non-linear operations. Given
%that many of the existing techniques commit to linear programs, they cannot
%handle this situation, although a linear ranking function does exist (see
%Figure~\ref{fig:handletable}).  In order to find a ranking function for
%this example, it is necessary to take into account the semantics of the
%bit-wise AND operator.
%
Our technique finds that an admissible ranking function is the linear
function $R(x) = x$, whose value decreases with every iteration, but cannot
decrease indefinitely as it is bounded from below.  This example also
illustrates the lack of a direct correlation between the linearity of a
program and that of its termination arguments.

\item Program (b) breaks assumption (ii), in that it has no linear ranking
function.  We prove that this loop terminates by finding the non-linear
ranking function $R(x) = |x|$.

@@

\item Program (d) also breaks assumption (iii), but ``the other way'': it
terminates for integers but not for bit-vectors.  If each of the variables
is stored in an unsigned $k$-bit word, the following entry state will lead
to an infinite loop:
%
$$ M = 2^k - 1,\quad N = 2^k - 1,\quad i = M,\quad j = N-1 $$

\item Program (e) breaks assumption (iv): it terminates for reals but not
for floats.  If $x$ is sufficiently large, rounding error will cause the
subtraction to have no effect.

\item Program (f) breaks assumption (iv) ``the other way'': it terminates
for floats but not for reals.  Eventually $x$ will become sufficiently small
that the nearest representable number is $0.0$, at which point it will be
rounded to $0.0$ and the loop will terminate.

\end{itemize}

Up until this point, we considered examples that are not soundly treated by
existing techniques as they don't fit in the range of programs addressed by
these techniques.  Next, we look at some programs that are handled by
existing termination tools via dedicated analyses.  We show that our method
handles them uniformly, without the need for any special treatment.
%
\begin{itemize}

\item Program (g) is a linear program that is shown
in~\cite{DBLP:conf/tacas/CookSZ13} not to admit (without prior manipulation)
a lexicographic linear ranking function.  With our technique we can find the
non-linear ranking function $R(x) = |x|$.

%As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
%loop transition leads to a decrease with respect to the lexicographic
%ordering for this tuple. Therefore no function may increase unless a function of
%a lower index decreases. Additionally, at every step, there must be at least one
%function that decreases.

@@

\item In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i)
admits a \emph{multiphase} ranking function, computed from a multiphase
ranking template.  Multiphase ranking templates are targeted at programs
that go through a finite number of phases in their execution.  Each phase is
ranked with an affine-linear function and the phase is considered to be
completed once this function becomes non-positive.

In our setting this type of programs does not need special treatment, as we
can find a non-linear lexicographic ranking function $R(x, y, z) = (x < y,
z)$.\footnote{This termination argument is somewhat subtle, but can be understood as
follows: The Boolean values $\mathit{false}$ and $\mathit{true}$ are
interpreted as 0 and 1, respectively.  The Boolean $x < y$ thus eventually
decreases, that is to say once a state with $x \geq y$ is reached, $x$ never
again becomes greater than $y$.  This means that as soon as the ``else''
branch of the if statement is taken, it will continue to be taken in each
subsequent iteration of the loop.  Meanwhile, if $x < y$ has not decreased
(i.e., we have stayed in the same branch of the ``if''), then $z$ does
decrease.  Since a Boolean only has two possible values, it cannot decrease
indefinitely.  Since $z > 0$ is a conjunct of the loop guard, $z$ cannot
decrease indefinitely, and so $R$ proves that the loop is well founded.}
\iffalse
\item Program (j) illustrates a nested loop.  This construction can cause difficulties for provers based on
enumerating lassos
\fi
\end{itemize}
%
As with all of the termination proofs presented in this paper, the ranking
functions above were all found completely automatically.

@@

\section{Preliminaries}

Given a program, we first formalise its termination argument as a ranking
function (Section~\ref{sec:ranking.functions}).  Subsequently, we discus
bit-vector semantics and illustrate differences between machine arithmetic
and integer arithmetic that show that the abstraction of bit vectors to
mathematical integers unsound (Section~\ref{sec:machine.arith}).

\subsection{Termination and Ranking Functions} \label{sec:ranking.functions}

A programs $P$ is represented as a transition systems with state space $X$,
transition relation $T \subseteq X \times X$, and set of variables $V$.  A
state of the program is a valuation of the variables from $V$.  For a state
$x \in X$ with $T(x,x')$ we say $x'$ is the successor of $x$ under $T$.

\begin{definition}[Unconditional termination]
%
A transition system is said to be \emph{unconditionally terminating} if
there is no infinite sequence of states $x_1, x_2, \ldots \in X$ with
$\forall i.~T(x_i, x_{i+1})$.
%
\end{definition}

We can prove that the transition system is unconditionally terminating by
finding a ranking function for its transition relation.
%
\begin{definition}[Ranking function]
%
An injective function ${R:X\to Y}$ is a \emph{ranking function} for the
transition relation $T$ if $Y$ is a well-founded set with order $<$ and 
$R$ is monotonically decreasing with respect to $T$.  That is
to say:
$$\forall x, x' \in X. T(x, x') \Rightarrow R(x) < R(x')$$
%
\end{definition}

@@

In the case that $\dim(Y) = 1$, this reduces to the inner product
%
$$f(\vec{x}) = \vec{\lambda} \cdotp \vec{x} + c \;.$$

@@

\subsection{Machine Arithmetic and Bit Vectors} \label{sec:machine.arith}

Physical computers have bounded storage, which means they are unable to
perform calculations on mathematical integers.  For example, if $A$ is the
Ackermann function and $G$ is Graham's number, a physical computer capable
of computing $A(G, G)$ would contain (much!) more matter than is believed to
exist in the universe.  Fortunately, it is rare for a programmer to need
such a large number and so modern computers do their arithmetic over
fixed-width binary words, otherwise known as bit vectors.  For the remainder
of this section, we will say that the bit vectors we are working with are
$k$-bits wide, which means that each word can hold one of $2^k$ bit
patterns.  Typical values for $k$ are 32 and 64.

Machine words can be interpreted as ``signed'' or ``unsigned'' values. 
Signed values can be negative, while unsigned values cannot.  The encoding
for signed values is two's complement, where the most significant bit
$b_{k-1}$ of the word is a ``sign'' bit, whose weight is $-(2^k - 1)$ rather
than $2^k - 1$.  Two's complement representation has the property that
$\forall x .  -x = (\mathord{\sim} x) + 1$, where $\mathord{\sim}(\bullet)$
is bitwise-negation.  Two's complement also has the property that addition,
multiplication and subtraction are defined identically for unsigned and
signed numbers.

Bit-vector arithmetic is performed modulo $2^k$, which is the source of many
of the differences between machine arithmetic and Peano
arithmetic.\footnote{ISO C requires that unsigned arithmetic is performed
modulo $2^k$, whereas the overflow case is undefined for signed arithmetic. 
In practice, the undefined behaviour is implemented just as if the
arithmetic had been unsigned}.  To give an example, $(2^k - 1) + 1 \equiv 0
\pmod {2^k}$ provides a counterexample to the statement $\forall x. 
x + 1 > x$, which is a theorem of Peano arithmetic but not of modular
arithmetic.  When an arithmetic operation has a result greater than $2^k$,
it is said to ``overflow''.  If an operation does not overflow, its
machine-arithmetic result is the same as the result of the same operation
performed on integers.

The final source of difference between integer arithmetic and bitvector
arithmetic stems from width conversions.  Many programming languages allow
numeric variables of different types, which can be represented using words
of different widths.  In C, a \texttt{short} might occupy 16 bits, while an
\texttt{int} might occupy 32 bits.  When a $k$-bit variable is assigned to a
$l$-bit variable with $l < k$, the result is truncated $\mod 2^l$.  For
example, if $x$ is a 32-bit variable and $y$ is a 16-bit variable, $y$ will
hold the value $0$ after the following code is executed:
%
\begin{lstlisting}
x = 65536;
y = x;
\end{lstlisting}

@@

As well as machine arithmetic differing from Peano arithmetic on the
operators they have in common, computers have several ``bitwise'' operations
that are not taken as primitive in the theory of integers.  These operations
include the Boolean operators \texttt{and, or, not, xor} applied to each
element of the bit vector.  Computer programs often make use of these
operators, which are non-linear when interpreted in the standard model of
Peano arithmetic\footnote{Some of these operators can be seen as
linear in a different algebraic structure, e.g.~\texttt{xor} corresponds to
addition in the Galois field $\mathrm{GF}(2^k)$.}.

@@

\begin{definition}[Synthesis Fragment]
A formula is in the \emph{synthesis fragment} iff it is of the form
%
 \[
  \exists \vec{P},~ \vec{x} . ~\forall~ \vec{y} . \sigma(\vec{P}, \vec{x}, \vec{y})
 \]
%
where $\vec{P}$ ranges over sets, while $\vec{x}$ and $\vec{y}$ range over
ground terms.  Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times
Y^m \to \mathbb{B}$ can be seen as a specification function that returns
true iff the functions $\vec{P}$ compute appropriate outputs when fed the
inputs $\vec{x}$ and $\vec{y}$.
%
\end{definition}
%
Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
%Solving the synthesis problem 
amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
for all $\vec{y}$. 
If a pair $(\vec{P_0}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P_0}, \vec{x}) \models \sigma$.
For the remainder of the presentation, we drop the $\vec{x}$ and instead just write $x$, with the understanding
that all variables range over vectors.

@@

We will begin our discussion by showing how to encode the
\mbox{(non-)termination} of a program consisting of a single loop in the
synthesis fragment.  For the time being, a loop is a pair ${L= \langle G,T
\rangle}$, where the states $x$ satisfying the loop's guard are given by the
predicate $G(x)$.  The body of the loop is encoded as the transition
relation $T(x, x')$, meaning that state $x'$ is reachable from state $x$ via
a single iteration of the loop body.  For example, the loop in
Figure~\ref{fig:motivation.a} is encoded as:
%
\begin{align*}
G(x) & = \{ x \mid x>0 \} \\
T(x,x') &= \{ \langle x, x' \rangle \mid x' = (x - 1) \, \& \, x \}
\end{align*}
We will abbreviate this with the notation:
\begin{align*}
G(x) & \triangleq x > 0 \\
T(x, x') & \triangleq x' = (x - 1) \, \& \, x
\end{align*}

@@

\begin{definition}[Non-Termination Formula -- Open Recurrence Set  {\bf [ONT]}]
\label{def:ont}
 \begin{align*}
  \exists N, x_0 . \forall x . \exists x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \rightarrow T_u(x, x') \wedge N(x') 
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula -- Closed Recurrence Set {\bf [CNT]}]
\label{def:cnt}
 \begin{align*}
  \exists N, x_0 . \forall x, x' . & N(x_0) ~ \wedge ~ N(x) \rightarrow G(x) ~ \wedge \\
							& N(x) \wedge T_u(x, x') \rightarrow N(x') 
 \end{align*}
\end{definition}

@@

As the existence of a ranking function is equivalent to the satisfiability
of the formula {\bf [UT]}, a satisfiability witness is also a ranking
function and thus a proof of $L$'s unconditional termination.

Continuing the example of the program in Figure~\ref{fig:motivation.a}, we
can see that the corresponding synthesis formula {\bf [UT]} is satisfiable,
as witnessed by the function $R(x) = x$.  Thus, $R(x) = x$ constitutes a
proof that the program in Figure~\ref{fig:motivation.a} is unconditionally
terminating.

Note that different formulations for unconditional termination are possible. 
We are aware of a proof rule based on transition invariants, i.e.  supersets
of the transition relation's transitive
closure~\cite{DBLP:conf/pldi/GrebenshchikovLPR12}.  This formulation assumes
that the second-order logic has a primitive predicate for disjunctive
well-foundedness.  Conversely, our formulation in Definition~\ref{def:UT}
does not require a transitive closure operator, nor a disjunctive
well-foundedness predicate.  \\

@@

Gupta et al.~\cite{DBLP:conf/popl/GuptaHMRX08} characterize non-termination
of a transition relation $T$ by the existence of an \emph{(open) recurrence
set}, i.e.  a nonempty set of states $N$ such that for each $s \in N$ there
exists a transition to some $s'\in N$.
For our case, the notion of open recurrent set is encoded by formula {\bf [ONT]}. %Definition~\ref{def:nonterm-formula}.
Dual to the termination problem, when formulating $L$'s non-termination, we only need to consider an under-approximation of its transition
relation~$T_u$. 
%$L = \langle G,T \rangle$
Again, this distinction is mostly important when dealing with nested loops,
as illustrated in Section~\ref{sec:env}.

@@

If this formula is satisfiable, $N$ is an open recurrence set for $L$, which proves
$L$'s non-termination. The issue with this formula is the additional level of quantifier alternation as compared to the synthesis fragment
(it is an $\exists \forall \exists$ formula). %This means that we will not be able to use the solver for the synthesis fragment to solve it.

In order to formulate the non-termination problem inside the synthesis
fragment, we make use of the notion of \emph{closed recurrence set}
introduced by Chen et al.~in~\cite{DBLP:conf/tacas/ChenCFNO14}.  As opposed
to an open recurrence set, this is an $\exists \forall$ property: for each
state in the recurrence set $N$, all of its successors must be in $N$.  We
note that if $T$ is deterministic, every open recurrence set is also a
closed recurrence set (since each state has at most one successor).  Thus,
the non-termination problem for deterministic transition systems is
equivalent to the satisfiability of formula {\bf [CNT]}.

%Definition~\ref{def:deterministic-nonterm-formula}.  As our formulation is
%with respect to a given loop $L=\langle I,G,T \rangle$, we require the
%recurrence set to include the loop's guard, $N(x) \rightarrow G(x)$.  In
%other words, the loop's guard needs to be satisfied in order for the
%infinite execution to happen.

@@

Note that by defining $T_u$'s determinism as the existence of exactly one
successor for any state satisfying $L$'s guard, we eliminate the possibility
of $T_u$ being $\mathit{false}$, which would otherwise constitute a
counterexample for Theorem~\ref{thm:cnt}.

Given that our program's state space is finite, a transition relation
induces a infinite execution if and only if some state is visited infinitely
often, or equivalently the state is visited twice in the course of the
execution.  This gives us an alternative way of expression a non-termination
proof: we can can try to find a witness $x$ to the satisfiability of the
formula
%
$$ \exists x . T^+(x, x)$$
%
However, deciding satisfiability of this formula directly requires a logic
that includes a transitive closure operator, which ours does not have.\\

\noindent {\bf Non-determinism.} A non-termination diagnosis procedure based
on checking the satisfiability of formula {\bf [CNT]} may fail to find a
closed recurrence set for a loop $L=(G, T_u)$ when $T_u$ is
non-deterministic.  To understand the issue, imagine the following loop with
a non-deterministic assignment to \lstinline!y!:
